6.867 Problem Set 2

### 2015-11-05 HW2/tex/HW2.tex
Final Draft

Table 3 shows the weights generated by our soft SVM algorithm, as well as its performance on the training and validation sets, given the four datasets we've been provided for testing. The soft SVM, in this case, uses a simple linear kernel, so the decision boundaries that are generated are still straight lines in $\mathbb{R}^2$. However, the slack provided by the soft SVM allows the algorithm to classify data and create a ``margin", even when the data is not linearly separable. As expected, in general the algorithm does slightly better on the training set than the validation set. Also, however, note that as the data becomes less and less linearly separable, our soft SVM with a linear kernel does increasingly poorly. This motivates our decision to explore the use of a Gaussian (or RBF) kernel.

Table 4 includes the performance of our soft SVM algorithm on various training sets when we change kernels (we provide data on both linear and Gaussian kernels), and also when we modify the bandwidth of our Gaussian kernel (where applicable) and change our regularization parameter, $C$. In general, our geometric margin increases as we decrease $C$. This is to be expected - as we provide the SVM with more slack, it is able to allow more points within its margin, and can consequently expand its geometric margin. In general, the larger the geometric margin, the more support vectors - again, this is intuitive. The more slack we allow, the more data points we will observe on or inside of our margin. It is not, however, necessarily true that the geometric margin will continue to increase as we decrease $C$ (and also, geometric margin will not always decrease as we increase $C$). As C gets very large, our decision boundary will approach the hard SVM decision boundary, and will not move at all. As $C$ gets increasingly small, the additional slack will lead our decision boundary to encompass all of the data, at which point our geometric margin will stop increasing.

@@

\begin{table}
\captionof{table}{Performance of Linear SVM on provided data sets}
\begin{tabular}{c c|c|c|c|c|c}
\toprule
{} & dataset & $w_0$ & $w_1$ & $w_2$ & classification error rate (training) & classification error rate (validation) \\
\midrule
  & data\_stdev1 & 0.31 & 5.42 & 0.87 & 0.00\% & 0.00\% \\
  & data\_stdev2 & 9.09e-2 & 9.83e1 & 9.11e1 & 9.50\% & 7.50\% \\
  & data\_stdev4 & -2.05e-2 & 2.58e2 & 2.40e2 & 26.00\% & 23.50\% \\
  & data\_nonsep & 2.56e-3 & 4.23e2 & 3.84e2 & 49.50\% & 51.25\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\captionof{table}{Soft SVM Performance for Different Parameters and Kernels} 
\centering
\begin{tabular}{llllll}
\toprule
{} &                        Kernel &      $C$ &   $\sigma$ &    Number of Support Vectors & $1/||\mathbf{w}||$ \\
\midrule
&    Gaussian &    .01 &    .1 &   400 &    .0286 \\
&    Gaussian &    .01 &    1 &   400 &    .0286 \\
&    Gaussian &    .01 &    10 &   400 &    .0286 \\
&    Gaussian &    .1 &    .1 &   400 &    .0141 \\
&    Gaussian &    .1 &    1 &   400 &    .0207 \\
&    Gaussian &    .1 &    10 &   400 &    .0141 \\
&    Gaussian &    1 &    .1 &   391 &    .0022 \\
&    Gaussian &    1 &    1 &   100 &    .0076 \\
&    Gaussian &    1 &    10 &   368 &    .0018 \\
&    Gaussian &    10 &    .1 &   390 &    .0018 \\
&    Gaussian &    10 &    1 &   71 &    .0017 \\
&    Gaussian &    10 &    10 &   209 &    .0003 \\
&    Gaussian &    100 &    .1 &   390 &    .0018 \\
&    Gaussian &    100 &    1 &   59 &    .0002 \\
&    Gaussian &    100 &    10 &   159 &    7.51e-5 \\
&    Linear &    .01 &    N/A &   399 &    .0282 \\
&    Linear &    .1 &    N/A &   393 &    .0133 \\
&    Linear &    1 &    N/A &   392 &    .0017 \\
&    Linear &    10 &    N/A &   393 &    .0002 \\
&    Linear &    100 &    N/A &   397 &    1.8e-5 \\
\midrule
\bottomrule
\end{tabular}
\end{table}

\subsubsection*{Titanic Data}
We're interesting in now testing the performance of our classifiers on a real dataset. In this case, we are using the Titanic dataset, where we're performing the rather macabre task of predicting passenger survival given his or her features. For this problem we first scale all the features so they lie between $[0,1]$. We accomplish this by taking each observation and transforming it via:
\begin{equation*}
	x_{i,j}^* = \frac{x_{i,j}-\min(x_j)}{\max(x_j) - \min(x_j)}
\end{equation*}
This scaling of our data ensures that logistic regression behaves nicely. Otherwise, we encounter precision loss errors that end up significantly impacting our predictive accuracy.

@@

To better understand the differences in our models, one can look at Figure 5, which compares the weights of different variables for our different models. Note that each model's array of weights is normalized so that its maximum weight (in this case, always gender) is equal to 1. One can see that the linear SVM places almost all of its weight on gender, with a small amount on \# of parents/children. Logistic regression also places the most weight on gender, but also puts significant weight on (in order of decreasing magnitude) 3rd class passengers, age, \# of parents and children, 1st class passengers, and 2nd class passengers. The polynomial SVM puts most weight on passenger class and passenger embarkation locations. And the Gaussian SVM, after age, places the most weight on 3rd class passengers, 2nd class passengers, and passengers who embarked in Cherbourg.

>>>>

Table 3 shows the weights generated by our soft SVM algorithm, as well as its performance on the training and validation sets, given the four datasets we've been provided for testing. The soft SVM, in this case, uses a simple linear kernel, so the decision boundaries that are generated are still straight lines in $\mathbb{R}^2$. However, the slack provided by the soft SVM allows the algorithm to classify data and create a ``margin", even when the data is not linearly separable. As expected, in general the algorithm does slightly better on the training set than the validation set. We plot our linear SVM classifier below:

However, note that as the data becomes less and less linearly separable, our soft SVM with a linear kernel does increasingly poorly. This motivates our decision to explore the use of a Gaussian (or RBF) kernel. Table 4 includes the performance of our soft SVM algorithm on various training sets when we change kernels (for both linear and Gaussian kernels), and also when we modify the bandwidth of our Gaussian kernel (where applicable) and change our regularization parameter, $C$.

In general, our geometric margin increases as we decrease $C$. This is to be expected - as we provide the SVM with more slack, it is able to allow more points within its margin, and can consequently expand its geometric margin. In general, the larger the geometric margin, the more support vectors - again, this is intuitive. The more slack we allow, the more data points we will observe on or inside of our margin. It is not, however, necessarily true that the geometric margin will continue to increase as we decrease $C$ (and also, geometric margin will not always decrease as we increase $C$). As C gets very large, our decision boundary will approach the hard SVM decision boundary, and will not move at all. As $C$ gets increasingly small, the additional slack will lead our decision boundary to encompass all of the data, at which point our geometric margin will stop increasing.

@@

\begin{table}
\centering
\captionof{table}{Classification Performance of Linear SVM}
\begin{tabular}{c c|c|c|c|c|c}
\toprule
{} & dataset & $w_0$ & $w_1$ & $w_2$ & Training Error & Validation Error \\
\midrule
  & stdev1 & 0.31 & 5.42 & 0.87 & 0.00\% & 0.00\% \\
  & stdev2 & 9.09e-2 & 9.83e1 & 9.11e1 & 9.50\% & 7.50\% \\
  & stdev4 & -2.05e-2 & 2.58e2 & 2.40e2 & 26.00\% & 23.50\% \\
  & nonsep & 2.56e-3 & 4.23e2 & 3.84e2 & 49.50\% & 51.25\% \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}
\captionof{table}{Soft SVM Performance for Different Parameters and Kernels} 
\centering
\begin{tabular}{llllll}
\toprule
{} &                        Kernel &      $C$ &   $\sigma$ &    \# SV & $1/||\mathbf{w}||$ \\
\midrule
&    Gaussian &    .01 &    .1 &   400 &    .0286 \\
&    Gaussian &    .01 &    1 &   400 &    .0286 \\
&    Gaussian &    .01 &    10 &   400 &    .0286 \\
&    Gaussian &    .1 &    .1 &   400 &    .0141 \\
&    Gaussian &    .1 &    1 &   400 &    .0207 \\
&    Gaussian &    .1 &    10 &   400 &    .0141 \\
&    Gaussian &    1 &    .1 &   391 &    .0022 \\
&    Gaussian &    1 &    1 &   100 &    .0076 \\
&    Gaussian &    1 &    10 &   368 &    .0018 \\
&    Gaussian &    10 &    .1 &   390 &    .0018 \\
&    Gaussian &    10 &    1 &   71 &    .0017 \\
&    Gaussian &    10 &    10 &   209 &    .0003 \\
&    Gaussian &    100 &    .1 &   390 &    .0018 \\
&    Gaussian &    100 &    1 &   59 &    .0002 \\
&    Gaussian &    100 &    10 &   159 &    7.51e-5 \\
&    Linear &    .01 &    N/A &   399 &    .0282 \\
&    Linear &    .1 &    N/A &   393 &    .0133 \\
&    Linear &    1 &    N/A &   392 &    .0017 \\
&    Linear &    10 &    N/A &   393 &    .0002 \\
&    Linear &    100 &    N/A &   397 &    1.8e-5 \\
\midrule
\bottomrule
\end{tabular}
\end{table}

\subsubsection*{Titanic Data}
We're interested in now testing the performance of our classifiers on a real dataset. In this case, we are using the Titanic dataset, where we're assigned with the rather macabre task of predicting passenger survival given his or her features. For this problem we first scale all the features so they lie between $[0,1]$. We accomplish this by taking each observation and transforming it via:
\begin{equation*}
	x_{i,j}^* = \frac{x_{i,j}-\min(x_j)}{\max(x_j) - \min(x_j)}
\end{equation*}
This scaling of our data ensures that logistic regression behaves nicely. Otherwise, we encounter precision loss errors that end up significantly impacting our predictive accuracy.

@@

To better understand the differences in our models, one can look at Table 6, which reports the calculated weights for each of our trained models. We also depict this visually in Figure 5. Note that in Figure 5, each model's array of weights is normalized so that its maximum weight (in this case, always gender) is equal to 1. One can see that the linear SVM places almost all of its weight on gender, with a small amount on \# of parents/children. Logistic regression also places the most weight on gender, but also puts significant weight on (in order of decreasing magnitude) 3rd class passengers, age, \# of parents and children, 1st class passengers, and 2nd class passengers. The polynomial SVM puts most weight on passenger class and passenger embarkation locations. And the Gaussian SVM, after age, places the most weight on 3rd class passengers, 2nd class passengers, and passengers who embarked in Cherbourg.
